\addcontentsline{toc}{chapter}{Introduction}
\chapter*{Introduction}

\hugered{O}N MAY 9, 1911, the Optina Elder, Hiero-schemamonk Joseph, fell asleep in the Lord. \label{intro-excerpt} He was the closest disciple of the great Elder, Hiero-schemamonk Amvrosy--closest not in external things only, but in spirit, in strength of obedience, devotion, and love. He was in truth “the beloved child” of the Elder Amvrosy, who looked upon him with hope, who rejoiced in him seeing his spiritual progress, and whom he in every way preserved and guarded with his fatherly hand from all corrupting winds of conceit, vainglory, and pride.

This child of obedience was reared within the walls of Elder Amvrosy's humble, lowly hut, which was permeated with the precepts of the great Elders Lev and Makary and with the prayers of their successor, the Elder Amvrosy. Here in this small cell, which became for him “the school of piety,” he undertook the study of what is in truth the highest of sciences, monasticism, and in time he also became a director of monastics. And how simple and unassuming all this was; by many it was not even noticed.

The character of Fr. Joseph was distinguished by exceptional modesty, sensitivity, and compliance; and with time, these qualities deeply penetrated his whole being and developed into the great virtues of humility, love, and angelic meekness.

Of those who knew Batiushka Joseph as a simple monk and cell-attendant of Elder Amvrosy, who does not remember his manner when he went at times to the women's side of the hut: his humble step, lowered eyes, short answer with a bow, his invariable modest smile of greeting, always inconspicuous, never obtrusive. Nonetheless, even then everyone almost instinctively turned to him with special reverence, for they perceived something in him which placed him above all others.

At that time, the Elder, Fr. Amvrosy, had a senior cell-attendant, Fr. Michael, who on entering the women's section of the hut, liked to speak a little and to say something instructive. Then the hut would become enlivened-they would ask various questions, and at times even soft, restrained laughter would be heard. However, when Fr. Joseph went, everything would become quiet, everyone would become serious, and many would trustingly communicate to him their griefs or a message for the Elder.