\addcontentsline{toc}{chapter}{The Kozelsk Monastery of the Entry of the Mother of God into the Temple}
\chapter*{The Kozelsk Monastery of the Entry of the Mother of God into the Temple\\ (Kozelskaya-Vvedenskaya)\footnote{A. A. Pavlovskiy, comp., \textit{A Comprehensive Illustrated Guide to the Monasteries and Holy Places of the Russian Empire and Athos} (Nizhny Novgorod, 1907), pp. 246-8.}\markboth{The Kozelsk Monastery}{}}

\hugered{T}HE KOZELSK Optina Hermitage of the Entry of the Mother of God into the Temple is located 4 versts (2 \sfrac{2}{3} miles) from the city of Kozelsk which lies on the Ryazan-Ural rail line. It is situated on the picturesque right bank of the Zhizdra River. According to tradition, this hermitage was founded prior to the fifteenth century by a certain repentant ataman (head) of a den of thieves, Opta, who likewise founded the Optin-Troitsky Monastery in the Orlov province.

In the fifteenth century monks and nuns lived adjacently in the monastery, but in 1499 the latter were taken to the city where they founded their own Convent of the Ascension. In the beginning the hermitage did not have great means, and in 1724 it was closed and turned into a parish, while the monks were transferred to Belev, to the Monastery of the Transfiguration of the Saviour.

At the intercession of local citizens, the Shepelevs and Ardatov, within two years after its closure the monastery was again restored at the expense of those who had interceded. In 1764 the hermitage became self-supporting, but apparently it again began to fall into decline, until it was restored by Archimandrite Makary in 1795. From then on, better times began for the hermitage, and it began to acquire great renown.

Among the number of holy things to be found in the Kazan winter church is a wonderworking icon of the Mother of God of Kazan, donated in 1808 by the landowner Soburovaya. In addition, there is a specially venerated icon of the Beheading of Saint John the Forerunner, in whose honor there is at the hermitage the Skete of Saint John the Forerunner. The Great Prince Constantine Constantinovich has here also donated a perpetually-burning icon lamp. The monastery's vestry is not large. The church feast is on the day of the Entry of the Mother of God.

The monastery has land in various provinces amounting to two thousand \textit{dessiatinas} (5,400 acres) of forest and agricultural land. There are about three hundred brethren under the direction of Archimandrite Xenophont.

\clearpage
\subtitle{
	Optina Hermitage\footnote{\textit{The Complete Orthodox Theological Encyclopedic Dictionary}, s.v. ``Kozelskaya-Vvedenskaya Optina.'' No author, editor, or date given (approximately 1910).}
}

Men's hermitage (self-supporting from 1764) in Kaluga Province, 4 versts from the city of Kozelsk on the river Zhizdra. The exact time of its founding is not known. The hermitage was destroyed by the Lithuanians and left desolate. It now has 6 stone churches, a hospice, guest-house, house, hospital. Eldership -- preserved in few monasteries -- is observed in the hermitage. Governed by an abbot. Approximately 310 brethren.
